---
title: "IDEAL Methylation Analysis Final"
author: "David Martino \nTelethon Kids Institute  \nUniversity of Western Australia\n"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    df_print: paged
  html_notebook:
    toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=8, fig.height=8)
```

**Project**
*Study: IDEAL project*  
*PI: Ofer Levy*  
*contact: Ofer Levy <Ofer.Levy@childrens.harvard.edu>*   

**Protocol**
*Cell type: PBMC*  
*Extraction method: Chemagic low volume blood kit*  
*Assay: Infinium Human Methylation EPIC array*  

**Data Generation**  
*0.65 - 1.5ug of DNA send to AGRF for bisulphite treatment and hybridization to array*  
*Contact: Melinda Ziino <melinda.ziino@agrf.org.au>*  

# Setup
## Directories
```{r}
datadir <- '/home/nschultz/file_workspace/Collaborative\ Projects/HPC_CEPI/ANALYSIS/CAGRF20083570_EPIC_IDEAL'
reportdir <- '~/file_workspace/Collaborative\ Projects/HPC_CEPI/ANALYSIS/CAGRF20083570_EPIC_IDEAL/results'
plotdir <- '~/file_workspace/Collaborative\ Projects/HPC_CEPI/ANALYSIS/CAGRF20083570_EPIC_IDEAL/plots'
```

## Libraries 
```{r call library packages, message= FALSE}
library(minfi)
library(ggplot2)
library(ggsci)
library(tidyverse)
library(ggpubr)
library(limma)
library(dplyr)
library(DMRcate)
library(regioneR)
library(gridExtra)
library(grid)
library(ggrepel)
```


```{r}
# using the PBS dmps

olprobes <- read.csv(file.path(reportdir,'IPstatusPBS.dmps.csv'))

Probes = olprobes$ProbeID

#subset to those probes
beta.sub <- Beta[Probes,]

#subset to PBS and MPL only
beta.sub <- beta.sub[,metadata$Condition=='PBS' | metadata$Condition=='MPL']

#also metadata
meta.sub <- metadata[metadata$Condition=='PBS' | metadata$Condition=='MPL',]

beta.sub.mpl <- beta.sub[, meta.sub$Condition=='MPL']
beta.sub.pbs <- beta.sub[, meta.sub$Condition=='PBS']

#calculate the delta of post - pre stimulation
#beta.delta <- beta.sub.mpl - beta.sub.pbs 
metadelta <- meta.sub[meta.sub$Condition=='MPL',] #subset to just one condition for linear modelling
```


```{r}
design <- model.matrix(~Ipstatus + sex + Age.months, metadelta) 

fit <- lmFit(beta.sub.mpl, design)
fit2 <- eBayes(fit)

result.mpl <- topTable(fit2, coef = 2, number = Inf, sort.by = 'none') 
```


```{r}
summary(result.mpl$adj.P.Val <= 0.05)

```


```{r}
fit <- lmFit(beta.sub.pbs, design) # Now we fit again to the pre-stimulation data set (PBS condition)
fit2 <- eBayes(fit)
result.pbs <- topTable(fit2, coef = 2, number = Inf, sort.by = 'none') # representes the statistical difference between groups pre stimulation
```


```{r}
summary(result.pbs$adj.P.Val <= 0.05)
 
```


```{r}
# subest to the 217 significant ones that were different at baseline in both results tables
result.pbs <- result.pbs[result.mpl$adj.P.Val >= 0.05,]
result.mpl <- result.mpl[rownames(result.pbs),]

# Calculate the sign of each element in each vector
pbs_sign <- sign(result.pbs$logFC)
mpl_sign <- sign(result.mpl$logFC)

# Compare the signs of the corresponding elements in the two vectors
signs_equal <- pbs_sign == mpl_sign

# Print the results
print(signs_equal)

#just get the ones where the signs match
result.mpl.sign <- result.mpl[signs_equal,]
result.pbs.sign <- result.pbs[signs_equal,]
```

#Bland Altman visualisation
```{r}
df = cbind(result.pbs$logFC, result.mpl$logF)
colnames(df)= c('effect.unstim', 'effect.MPL.stim')
rownames(df) = rownames(result.pbs)
df = as.data.frame(df)
```

```{r}
#create new column for average measurement
df$avg <- rowMeans(beta.sub[rownames(df),]) 

#create new column for difference in measurements
df$diff <- df$effect.MPL.stim - df$effect.unstim

#find average difference
mean_diff <- mean(df$diff)

#find lower 95% confidence interval limits
lower <- mean_diff - 1.96*sd(df$diff)

#find upper 95% confidence interval limits
upper <- mean_diff + 1.96*sd(df$diff)
```

```{r}
png(file=file.path(plotdir,'bland_atlman.png'), width = 650, height = 450)

#create Bland-Altman plot
ggplot(df, aes(x = avg, y = diff)) +
  geom_point(size=3, shape = 21,colour = "black",fill = '#8080ff') +
  geom_hline(yintercept = mean_diff) +
  geom_hline(yintercept = lower, color = "red", linetype="dashed") +
  geom_hline(yintercept = upper, color = "red", linetype="dashed") +
  ggtitle("Bland-Altman Plot") +
  ylab("Difference Between Effects") +
  xlab("Average Methylation")+
  theme_bw(base_rect_size = 2, base_size = 20) + 
   theme(plot.title = element_text(hjust = 0.5), axis.text = element_text(face = "bold"))
```
#Session Info
```{r}
sessionInfo()
```